Extended Early Adulthood and Attitude Change

### 2024-07-29 manuscript/manuscript.Rmd

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences. This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect biological or cognitive differences between young adults and older adults. In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues. Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s. Instead, we find that other age groups have become more stable opinion holders. These aggregate trends obscure two countervailing patterns: decreases in stability due to life-course changes among younger adults and increases in stability across all age groups due to increases in educational attainment, as well as a broader stability trend that remains unexplained. Our results suggest a braod role in social contexts in shaping attitude stability."
---

@@

@@

Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991]. In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].

Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Related explanations conversely focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

It is likely true that both social and cognitive/biological mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution. If the mechanisms are principally cognitive and biological, rather than social, then broader changes in the structure of different life stages are unlikely to affect rates of social change in the same way. If, conversely, openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].

Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005]. Rates of high school and college completion have also increased significantly. As a result, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

In this paper, we test whether these broader shifts in social structure and the timing of life course transitions have affected the rate at which young people change their attitudes about political and social issues. Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we estimate the amount of change in opinions as a function of age and cohort. Consistent with the "impressionable years" hypothesis, we find that people in age ranges associated with this "early adulthood" life stage (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as more recent cohorts exhibited more stable opinions in middle and old age compared to their predecessors, while early adult respondents' reported opinions continued to change at roughly the same rate as in earlier cohorts.

Furthermore, we show that the aggregate patterns in attitude change over time are principally due to forces other than changes in life-course structure. While decreased rates of marriage explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased rates of education across all age groups. Furthermore, much of the shift in aggregate attitude stability remains unexplained, suggesting broader social structural features such as increased access to information with which to form opinions, political and cultural polarization, and more are the key drivers of attitude stability.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level. In political socialization, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@ryder; @fosse2023a; @elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These aggregate models of cultural imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. At a basic level, the "impressionable years" hypothesis requires that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@]. These tests tend to find higher rates of attitude change among the youngest adults -- people ages 18-25 -- and little difference between adults of other age ranges. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains more variance than period, the year of observation, for most items in the General Social Survey, suggesting an impressionable-years pattern.

In recent years, researchers have also developed new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns to previous work with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019]. However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015 -- which has not been tested.

@@

While the empirical pattern of early openness followed by relative stability appears quite robust, its cause is not clear. At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both these processes change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

## Biological and Cognitive Mechanisms

The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and theories about how people process information. This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate [@bartels2014; @gopnik2020; @sherratt2018]. However, because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014]. From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

It is also possible that there is something unique about human cognition during these periods that makes people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015]. Supporting this model is evidence that early formative periods have been observed not just in people, but across species. For example, in a study of social learning among birds, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

In light of empirical evidence about these sensitive windows across species, there has been modeling work about why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015). The central argument from this work is that that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment, and that early formative periods are likely to evolve when environments are unstable and information is plentiful yet unreliable (Frankenhuis and Walasek 2020). In these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past. Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021). For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with at best contradictory information about what one should or should not like, or should or should not wear. Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.

The thrust of the biological and cognitive arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages, regardless of their social environments. However, while biological mechanisms can shape how people respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change. For that, we must turn attention to the structuring of the social world.

@@

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts. Here we briefly consider some of the ways that the "impressionable years" differ from other life stages that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.

First, people in this age group tend to move across social contexts at a much higher rates than people in other age groups. In the 11-year window from ages 14 to 25, in which people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information. People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups [NEED CITES HERE].

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions. For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021]. Personal social networks also differ in this time period. Young peoples' networks tend to be larger and include more friends than kin [@marsden1987]. It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

Third, these are periods are also culturally distinct, with exploration and self-discovery expected and encouraged. People in this "early adulthood" phase are not expected to be stable adults, something should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2007; @elder1994]. Cultural expectations to engage in new experiences, meet new people, and adopt new personalities could motivate the pursuit of new information during this life stage apart from social structural opportunities.

## Updating Conditional on Exposure

While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on people's attitudes is contingent on their ability to incorporate it into their cultural repertoire. The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.

Social networks not only shape peoples' exposure to new information, but they can also affect the degree to which people adopt new beliefs and attitudes. People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs, even when they are exposed to new ideas [@stark1980]. Consistent with this, mid-life is associated with denser, more homogenous networks that include larger numbers of family [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004]. If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks. Similarly, if information flows less freely through weak-tie networks, these structures might allow respondents to limit their exposure to or close off parts of their lives to people who might not be supportive of new attitudes.

Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power, which tend to occur more frequently during middle age. As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information or forcing them to develop stable opinions on issues where previously they felt capable of ambivalence [@eaton2009]. This potentially explains another phenomenon: that people appear to open up to changing their minds on some issues as they enter later life and retirement, when these social expectations of stability presumably diminish [@danigelis2007].

Existing research suggests that the structure of social life during adolesence, early adulthood, and later life stages shape both exposure to new information and a person's ability to update attitudes, opinions, and beliefs in the face of that information. This further suggests that changes in these life course stages should matter for the amount of opinion change we observe.

@@

In the United States, the last seven decades have seen considerable social and cultural changes around adolescence and young adulthood [@buchmann1989; @rosenfeld2007; @rosenfeld2005]. The expansion of higher education and increased possibilities for mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow [@rosenfeld2007]. These changes have made life course trajectories less predictable, and the "script" of young adulthood has become less rigid: it might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner [@buchmann1989]. For the purposes of this paper, there are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.

@@

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for successive cohorts as they pass through the same ages. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce. In the United States, both the median age at first marriage and median age of first childbirth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.

Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce or movement back into education after several years in the workforce. These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.

If the distinctive nature of "impressionable years" is principally driven by their distinct social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect on how long exploration, and therefore openness to change, lasts throughout the life-course. If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to older adults in their rates attitude change but that subsequent cohorts will diverge from older adults, making more change over time.

*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 18 to 25 than cohorts born in earlier years.*

We expect a similar but more muted trajectory for the attitude behavior of people aged 25-33. Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research [@krosnic1991; @alwin1991], but that more recent cohorts will come to more closely resemble the 18-25 age group, diverging from middle aged respondents.

@@

Consistent with life-course structuring theories, we expect these divergences will principally be explained by changes in the timing of life course transitions such as marriage and childbirth. Controlling for these life-course transitions should eliminate the gap between the youngest adults and later adults.

*Hypothesis 3: Controlling for rates of marriage and childbirth, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*

@@

: Table \@ref(tab:2): Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.

>>>>

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: 'Early adulthood is consistently identified as a "formative period" in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences. This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect ontogenetic differences between young adults and older adults. In this paper, we draw on data for more than 70 questions that were asked repeatedly across multiple panel surveys stretching from the 1950s to the 2020s to explore whether changes in the social conditions of early adulthood have affected the rate at which early adults and the rest of the population change their opinions on political and social issues. Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the late 1950s. Instead, we find that other age groups have become more stable opinion holders over this time. This aggregate trend reflects three countervailing patterns: small decreases in stability due to life-course changes among younger adults, larger increases in stability across all age groups due to increased educational attainment, and a general increase in stability that remains unexplained. Our results suggest broad roles for both individual and larger social contexts in shaping when in the life course people experience attitude change.'
---

@@

\doublespace

@@

Adolescence and early adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making at later life stages [@krosnick1989; @alwin1991]. In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@kiley2020; @krosnick1989], and that opinions formed in these age ranges tend to persist well into adulthood [@sears2013; @sears1999]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952; @voas2016].

Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear. Social-science explanations tend to argue that the unique social features of different life stages shape whether people are open to or closed off from change during these ages. These features include high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" in early adulthood [@alwin1991; @danigelis2007] and normative expectations around stability for middle-aged people in high-status roles [@eaton2009; @visser2004]. At the same time, interdisciplinary work on learning across the life-course finds that an early period of active exploration followed by relative stability is a broader phenomenon across organisms, which might be an evolutionary response to the tension between learning and using information in high-information environments [@gopnik2020; @frankenhuis2020; @spreng2019]. These findings suggest that the differences between early adults and the rest of the population might be more rooted in ontogenetic mechanisms than social mechanisms.

It is likely true that both social and ontogenetic mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but clarifying their relative contributions is important for understanding the broader relationship between social change and cultural evolution. If the mechanisms driving heightened rates of opinion change early in adulthood are principally ontogenetic, then changes in the structure of different life stages would be unlikely to affect rates of cultural evolution. If openness to belief and opinion change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing [@cherlin2004; @cherlin2021; @robbins2022], could shape the relative importance of within-person change and cohort effects for explaining broader cultural change [@kiley2020; @vaisey2016].

To better understand the mechanisms that drive openness to cultural change, we take advantage of changes in the social conditions of life stages over time. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, establishing an independent household, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts than they did for earlier cohorts [@buchmann1989; @bruckner2005]. As a result, "early adulthood" has emerged as a distinct life stage, in which people are relatively detached from family and other role obligations and have more freedom to explore ideas and lifestyles [@rosenfeld2009]. This extension of the social conditions typically thought to open people up to change should be reflected in how much people change over time.

In this paper, we test whether these shifts in social structure have affected the rates at which people change their opinions about diverse social and political issues during early adulthood. Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we quantify broad trends in attitude change and stability over time. Consistent with the "impressionable years'' hypothesis, we find that early adults (aged 18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant across cohorts. They emerged over the second half of the 20th Century because more recent cohorts became more stable opinion holders in middle-age than their predecessors, not because they demonstrated more opinion change in early adulthood than their predecessors.

We show that changes in the timing of life-course transitions do not appear to explain these aggregate patterns of attitude change over time. While decreased rates of marriage in early adulthood explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased levels of educational attainment across all age groups. At the same time, much of the increase in attitude stability over time remains unexplained, suggesting broader social structural features, such as increased access to information with which to form opinions or political and cultural polarization, might also be the key drivers of attitude stability.

@@

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @elder1974; @guhin2021]. These frameworks argue that the circumstances in which people are brought up leave an enduring mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level. In theories of political socialization, for example, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than political preferences. The often-invoked concept of cohort effects rests on the assumption that peoples' early years play a key role in shaping a wide variety of tastes, preferences, habits, and dispositions, which then impacts their choices and life outcomes well beyond the time when this period has passed [@ryder1965; @fosse2023a; @elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding social theories of behavior.

Evidence for the impressionable years hypothesis exists at the individual and aggregate levels. At the individual level, the "impressionable years" hypothesis requires that people demonstrate opinion change early in life followed by a longer period of relative stability, especially in middle age. While different attitudes occasionally follow different patterns, this general trend has been shown to be true for political attitudes including ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; and other dispositions such as personality traits [@wagner2019] and cultural tastes [@mulder2010; @ma2021]. More generally, recent research has found that durable change of social opinion is quite rare among U.S. adults [@kiley2020]. At the aggregate level, cohort effects are consistently strong predictors of a wide variety of social and political attitudes, and Vaisey and Lizardo [-@vaisey2016] find that the year of a person's birth explains more variance in opinion differences than the year of observation for most items in the General Social Survey, suggesting an impressionable-years pattern.

In recent years, researchers have developed new approaches to quantify openness to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns to previous work, finding that openness peaks from a person's mid-teens through age 25 and declines quickly after that, lending more credence to the "impressionable years" hypothesis (a pattern that Way [-@way2019] suggests also holds for music tastes). However, this work tends to assume a constant age-based rate of openness across cohorts – i.e. that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015. This assumption remains untested. Regardless, there is broad evidence for a consistent pattern of malleability early in life followed by relative stability.

@@

While the empirical pattern of early openness followed by relative stability appears robust, its causes are not clear. Changing political, social, and cultural attitudes involve at least two distinct processes that can be difficult to disentangle in observational data: exposure to new information and change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both these processes change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

## Ontogenetic Mechanisms

One class of explanations for heightened rates of attitude change in early adulthood posits that attitudinal stability and change across the life course is primarily a function of humans’ cognitive architecture, which remains more or less stable across socio-historical contexts. One class theories posit that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally and are exposed to information at a (more or less) constant rate as they age [@bartels2014; @gopnik2020; @sherratt2018]. Because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014]. From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

A second theoretical framework suggests that development of the human mind changes across the life course and that there is something unique about human cognition during early periods that makes humans ascribe more weight to the information acquired within them than to information acquired at other points in the life course. Maturational arguments suggest that animal brains exhibit more malleability during childhood and adolescence than later in life [@mata2015, Spreng and Turner], and researchers observe evidence of this increased malleability in a wide variety of species. Younger birds, for example, are able to retool their habits in light of new information while older birds were not [@aplin2017]. Evolutionary theories posit that this pattern is a response to conditions in which environments are unstable and information is plentiful yet unreliable (Fawcett and Frankenhuis 2015; Frankenhuis and Walasek 2020; Walasek, Frankenhuis, and Panchanathan 2021), three circumstances common in human social environments. Simulations show that in these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past.

The thrust of these ontogenetic arguments is that humans have evolved a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages regardless of their contemporaneous social conditions. However, while these mechanisms can shape how humans respond to new information when they encounter it, they cannot dictate the rate at which people are exposed to new information, which is a fundamentally social process.

@@

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct. There are at least four dimensions on which early adulthood differs from other life stages in ways that could produce heightened rates of opinion change, regardless of ontogenetic factors.

First, people in these life stages tend to move across social contexts at a higher rate than people in other age groups. In the 11-year window from ages 14 to 25 in which existing work suggests people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information and restructuring core discussion networks [@small2015]. People in this age range also show the highest rates of geographic mobility [@geist2008]. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility might explain part of the difference in attitude stability between age groups [@billari2010]. Social, contextual, and geographic mobility might also free people to adopt new opinions as they move to new contexts and are unburdened by previous friends’ or family members’ expectations.

Second, in addition to inhabiting a greater diversity of social contexts during these years, people in adolescence and early adulthood spend their time in social contexts explicitly designed to affect attitudes and opinions. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts expose young adults to new information in ways that can lead to new opinions. For example, college attendance appears to change some opinions about civil liberties, politics, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021].

Third, the structure of personal social networks is distinct in this age range, with young peoples' networks tending to be larger and including more friends than kin compared to networks in middle age [@marsden1987]. Having a higher number of "weak ties" in their networks can expose younger adults to new information at higher rates [@granovetter1973; @burt2005]. And because people are less likely to adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs [@stark1980], the denser, more homogenous networks that characterize middle-age [@marsden1987; @diprete2011] might make it more difficult for people in this life stage to update opinions even if they receive new information [@visser2004]. Conversely, young peoples’ less-dense networks might allow them to close off parts of their lives to people who might not be supportive of new attitudes.

Fourth, there is significant social variation in norms around when maturity is reached [@cunningham2007; @elder1994], and contemporary early adults have little cultural pressure to resemble middle-aged adults. Cultural narratives encourage exploration and self-discovery during early adulthood, pushing young adults to pursue new experiences, meet new people, and adopt new personalities during this “age of independence” [@rosenfeld2009]. These cultural expectations might also shape whether people feel capable of updating opinions in the face of new information. People appear to value stability in people holding social roles with authority [@eaton2009]. As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information. This potentially explains why some work finds that people appear to open up to changing their minds as they enter later life, when these social expectations of stability potentially diminish [@danigelis2007].

As the above suggests, life stages vary in myriad ways that could affect how people are exposed to new information and whether they can act on that information in ways that produce changes of opinion. As a result, changes in the duration of these life stages could have significant ramifications for how much opinion change we see in a population.

@@

The social and cultural conditions of early adulthood have changed considerably in the past seven decades in the United States [@buchmann1989; @rosenfeld2009; @rosenfeld2005], as expansion of higher education and increased possibility for geographic mobility have reshaped expectations around life-course trajectories [@rosenfeld2009]. While cultural norms previously encouraged marriage, stable employment, and childbearing shortly after the completion of secondary education, relaxation of these norms means that early adulthood now might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner [@buchmann1989, @billari2010]. There are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.

@@

The period between adolescence and "settling down" has gotten longer over time [@billari2010]. Buchmann [-@buchmann1989] finds that young adults increasingly postpone the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce. In the United States, the median age at first marriage and median age of first childbirth both occur about eight years later than they did in the mid-20th Century [@rosenfeld2009], and rates of non-marital childbearing have also increased across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts than they did for previous cohorts.

Similarly, more recent cohorts spend more years pursuing education, with a larger proportion of the U.S. population completing high school, college, and, increasingly, graduate degrees before entering the workforce [@schofer2005]. This shift has been particularly dramatic for women, whose educational attainment and labor-force participation are much higher than 70 years ago [@england2010]. People now also move between the workforce and education at higher rates than they did in the past [@rosenfeld2009].

If the heightened rates of attitude change during peoples’ "impressionable years" is principally driven by the distinct social circumstances of adolescence and early adulthood, then the delay of full adulthood has the potential to extend these social circumstances into later ages, extending heightened rates of attitude change later in the life course. While people born in the mid-20th century might have displayed mainly opinion stability by the time they reached their mid-twenties, when they married and had children, people from more contemporary cohorts, unmarried and childless at the same ages, might show higher rates of opinion change. Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to contemporaneous middle-aged adults in their rates of attitude change, but that these age groups will diverge in their rates of change in more recent years. This leads us to our first hypothesis:

*Hypothesis 1: More recent cohorts will show higher rates of opinion change in the age range of 18 to 25 than cohorts born in earlier years.*

We focus on people 18 to 25 for two reasons. First, existing work has explored attitude stability and change among people in this age range in earlier cohorts [@krosnick1991; @alwin1991], providing a baseline for what to expect. Second, this group has seen the most significant change in life course transitions of all age groups. Whereas most people in earlier cohorts were married with children by the midpoint of this age range, the median age of marriage and first childbirth is now outside this group for both men and women. We expect a similar but more muted trajectory for attitude behavior of people aged 25-33. Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research, but that more recent cohorts will come to more closely resemble the 18-25 age group. This produces our second hypothesis:

@@

If these differences are principally driven by the structure of the life course at these stages, then we would expect that, controlling for these life-course transitions, the gap between early adults and older adults will disappear. This leads to our third hypothesis:

*Hypothesis 3: Controlling for rates of marriage, childbirth, and having a child at home, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*

@@

: Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.
